---
title: "Data cleaning with tidyr, strings and lubridate"
author: "Liz Lucas, IRE"
output: 
---

Load your libraries! Since we'll use multiple tidyverse packages for this script we'll just load `tidyverse`. We'll also use `lubridate` to work with dates. `lubridate` must be installed and loaded separately though it is adjacent to the `tidyverse` world.

```{r}
library(tidyverse)
library(lubridate)
```

Load your data. We'll use three different files with different cleaning / reshaping needs:

-   Load disney_movies_total_gross.csv into a variable called `disney`
-   Load ohsa.csv into a variable called `osha`
-   Load poverty_original.csv into a variable called `poverty`

```{r}
disney <- read_csv("data/disney_movies_total_gross.csv")
osha <- read_csv("data/osha.csv")
poverty <- read_csv("data/poverty_original.csv")
```

Start with the `disney` file.

```{r}
glimpse(disney)
```

There are a bunch of issues with the formatting in this dataset: the dates are imported as text because they're written out, and the `total_gross` and `inflation_adjusted_gross` are both appearing as numbers because they were imported with dollar signs and commas. To start, let's deal with dates. `lubridate` offers many functions to convert dates formatted in different types. Check the documentation for different functions to address different formats; the one we'll use here is the `mdy()` function. 

The other function we'll use is `mutate()` which creates a new column and populates it with a value, formula or function. It's a lot like tacking a column onto the end of your data table in Excel or Google Sheets. 

```{r}
disney %>% 
  mutate(new_dt = mdy(release_date))

# I like to tack on a select() to easily see whether the date field converted properly:
disney %>% 
  mutate(new_dt = mdy(release_date)) %>% 
  select(new_dt, release_date)
```

See, it works! Let's go ahead and make that new column permanent, and use the `.after` argument to place the column in the middle of our table:

```{r}
disney <- disney %>% mutate(new_dt = mdy(release_date), .after=release_date)
```

Now to the numbers columns: `total_gross` and `inflation_adjusted_gross`. There's a function in `tidyverse` that comes from the `readr` package called `parse_number()` that will do all the work for you. See it in action:

```{r}
disney %>% mutate(new_total_gross = parse_number(total_gross)) %>% select(total_gross, new_total_gross)
```

👉YOUR TURN
Write some code to create two new columns to convert both `total_gross` and `inflation_adjusted_gross`:
```{r}

```

👉YOUR TURN
Now you you can ask, in inflation-adjusted dollars, which disney movie is the highest-grossing movie of all time?
```{r}

```

### Adding description columns

Next we'll take a look at `osha`.

```{r}
glimpse(osha)
```

If you look through the [documentation](https://www.osha.gov/sites/default/files/ITA_Data_Dictionary.pdf) for this dataset, you'll notice that some of these fields are coded, such as `size` and `establishment_type`. For columns that have many value options, we might want to join to a lookup table. But for just a few values, we can add a `_desc` column into our data and code in values based on the original column.

We'll add an `estab_type_desc` column based on the values in `establishment_type`, using a function called `case_when()`. This is something like an `if` or `ifelse` statement:

```{r}
# test it out
osha %>% mutate(estab_type_desc = case_when(
  establishment_type==1 ~ "Not a Government Entity",
  establishment_type==2 ~ "State Government Entity",
  establishment_type==3 ~ "Local Government Entity",
  TRUE ~ "Error"
)) %>% 
  count(establishment_type, estab_type_desc)

# make it permanent
osha <- osha %>% mutate(estab_type_desc = case_when(
  establishment_type==1 ~ "Not a Government Entity",
  establishment_type==2 ~ "State Government Entity",
  establishment_type==3 ~ "Local Government Entity",
  TRUE ~ "Error"
))
```


### Working with strings

For this exercise, we'll use `poverty_original`, which is the original file that I got from the Census Bureau. I cleaned it up using several techniques that I'll share here.

```{r}
glimpse(poverty)
```

First, let's do some string manipulation. There are many many things you can do to manipulate strings with the `stringr` package (included in `tidyverse`). [See the cheat sheet](https://github.com/ireapps/R-for-data-analysis-2022/blob/main/docs/strings.pdf).

I'll highlight a few. First, take a look at the `GEOID` column. This is a FIPS code column, standard in Census datasets, but it's a universal numbering system that identifies different geographies. The first two numbers of this FIPS code is the state FIPS, and the last three are the county FIPS. You can use these to easily join tables with other data that includes FIPS, without having to worry about whether the states and counties are spelled or represented exactly the same way.

Start by pulling out the first two numbers for the state FIPS, using the `str_sub()` function, which is one function you can use to subset strings:

```{r}
?str_sub

# arguments: (string to parse, start position, stop position)

str_sub(poverty$GEOID, 1, 2)

# test it out with mutate()
poverty %>% mutate(state_FIPS = str_sub(GEOID,1,2), .after=GEOID)

# create a new column
poverty <- poverty %>% mutate(state_FIPS = str_sub(GEOID,1,2), .after=GEOID)
```

Now try it out on the county FIPS:

```{r}
# test it out with mutate()
poverty %>% mutate(county_FIPS = str_sub(GEOID,3,5), .after=GEOID)

# create a new column
poverty <- poverty %>% mutate(county_FIPS = str_sub(GEOID,3,5), .after=GEOID)

```

Great. One last string to manipulate here, and that is the `NAME` column. In the data we worked with eariler this week, I had separated out county and state names to make it easier to work with. I did this using the `separate()` function, which allows you to split a column into multiple columns based on a delimiter:

```{r}
?separate
```

Note that you can identify a separator with the `sep =` argument; you can decide whether to remove the original column with the `remove =` argument (the default is TRUE).

```{r}
# test it out
poverty %>% separate(NAME, into=c("county","state"), sep=", ")

# try it with 'remove = FALSE'
poverty %>% separate(NAME, into=c("county","state"), sep=", ", remove = FALSE)
```

Now make it permanent

```{r}
poverty <- poverty %>% separate(NAME, into=c("county","state"), sep=", ", remove = FALSE)
```

There are many things you can do with `stringr`. Here are a few more examples:

```{r}
# join strings together
poverty %>% mutate(county_state = str_c(county, ", ", state)) %>% select(county, state, county_state)

# make all upper case
poverty %>% mutate(COUNTY = str_to_upper(county)) %>% select(county, COUNTY)

# make all lower case
poverty %>% mutate(county_lower = str_to_lower(county)) %>% select(county, county_lower)

# replace all instances of a string (in this case you should keep in mind that there are Parishes and Cities and Municipios as well)
poverty %>% mutate(new_county = str_replace(county, " County", "")) %>% select(county, new_county)

```

### Reshaping data

There's one last thing to do to the `poverty` table. In this example, each variable (i.e. `below50`, `below125`, etc) is its own row. To make it easier to do calculations by county, I transposed this data so that each variable would be its own column rather than row. I did that using `pivot_wider()` (for the sake of this example, I'm going to exclude the margin of error, or `moe`).

```{r}
poverty %>% 
  select(-moe) %>% 
  pivot_wider(names_from=variable, values_from=estimate)
```

There is a function called `pivot_longer()` that does the opposite:

```{r}
# First I'll create a new variable with the wider data:
poverty_wide <- poverty %>% 
  select(-moe) %>% 
  pivot_wider(names_from=variable, values_from=estimate)

# Then I'll turn it back to long using pivot_longer()
poverty_wide %>% pivot_longer(cols = population:below500, names_to="variable", values_to="estimate")

```
