---
title: "Practice with ggplot"
author: "Liz Lucas, IRE"
output:
---

```{r}
library(tidyverse)
library(lubridate)
library(forcats)
```

Three different datasets to illustrate various graphic options
```{r}
dispatch <- read_csv("data/dispatch.csv")
mlb <- read_csv("data/mlb.csv")
opioids <- read_csv("data/opioid-deaths.csv")
```

We'll start by trying out some different geoms:

### One variable (continuous)

**MLB**

```{r}
ggplot(mlb, aes(salary)) +
  geom_area(stat="bin")
```

```{r}
ggplot(mlb, aes(salary)) +
  geom_area(stat="bin", bins = 40)
```

```{r}
ggplot(mlb, aes(salary)) +
  geom_histogram()
```

```{r}
ggplot(mlb, aes(salary)) +
  geom_histogram(bins = 10, color = 1, fill = 2) +
  scale_x_continuous(labels=scales::dollar_format())
```

```{r}
ggplot(mlb, aes(salary)) +
  geom_histogram(bins = 10, color = 1, fill = 2) +
  scale_x_continuous(labels=scales::dollar_format()) +
  facet_wrap(~team)

```

### One variable (discrete)

**dispatch**

```{r}
dispatch %>% ggplot(aes(x=DOW)) +
  geom_bar()

# in order for the days of the week to appear in the right order, we need to turn that column into a factor:
dispatch %>% mutate(DOW = fct_relevel(DOW,
                                      "Sunday", "Monday", "Tuesday", "Wednesday", 
                                      "Thursday", "Friday", "Saturday")) %>% 
  ggplot(aes(x=DOW)) +
  geom_bar()
```

### Two variables (both continuous)

**opioids**

```{r}
# first look at the unaggregated dataset:
opioids %>% ggplot(aes(x=year, y=deaths)) +
  geom_point()

# might be easier to see total deaths by year and then graph it:
d1 <- opioids %>% group_by(year) %>% summarise(deaths = sum(deaths, na.rm=T))

d1 %>% ggplot(aes(x=year, y=deaths)) +
  geom_point()

# or you could use something like geom_smooth():
opioids %>% ggplot(aes(x=year, y=deaths)) +
  geom_smooth()

```

What we really want to do is look at the age adjusted rate, but we can't aggregate over it, but we can look at one state at a time:

```{r}
opioids %>%
  filter(state=="Missouri") %>% 
  ggplot(aes(x=year, y=age_adjusted_rate)) +
  geom_smooth()
```

Or use facets to compare all states:

```{r}
opioids %>%
  ggplot(aes(x=year, y=age_adjusted_rate)) +
  geom_smooth() +
  facet_wrap(~state)
```

### Two variables (one continuous one discreet)

**opioids**

```{r}
# create the dataset you need (age-adjusted death rate for all states in 2020)
d2 <- opioids %>% filter(year==2020) %>% drop_na(age_adjusted_rate)

d2 %>% 
  ggplot(aes(x=state,y=age_adjusted_rate)) +
  geom_col()
```

Let's try that again but reordering the states and changing the orientation of the labels:

```{r}
d2 %>% 
  mutate(state = fct_reorder(state, age_adjusted_rate)) %>% 
  ggplot(aes(x=state,y=age_adjusted_rate)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

### Do your own thing

```{r}

 
```
