---
title: "NICAR 2023: Intro to R and RStudio"
author: "Ryann Grochowski Jones"
date: '2023-03-04'
output: html_document
---

*In this hour-long session, we'll explore how to sort, filter and group by using R in RStudio. This notebook is adapted from the "R 101" tutorial designed by the folks at IRE-NICAR.*

First off, load the tidyverse set of packages.
Run the following code chunk by hitting the green play button in the upper-right corner.

```{r}
library(tidyverse)
```

Next up, load some data into our environment to play with.

We'll spend some time later learning how to use read functions to bring a variety of data types into RStudio.

But for now, let's load another package, called babynames, that has a nice and tidy data set ready for us to use.

```{r}
library(babynames)
```

Now run this next piece of code to bring the data set into our environment.

You do this by storing the data in a variable that we'll call "babynames_data."

```{r}
babynames_data <- babynames
```

In your environment pane (upper right), see how "babynames_data" appeared under a "Data" heading in our environment?

Click on "babynames_data" and it will show you the table in a new tab (next to the name of this script).

Now that we've got our data loaded into a variable, we can refer to the data using that variable name.
Next we'll use a pipe (`%>%`) to run the data through a function.
The first function we'll try out is `arrange()`, which sorts data.
Arrange your data by number of names to see what names were used the least amount of times.

```{r}
babynames_data %>% arrange(n)
```

A bunch of names were only used 5 times in a year (the minimum amount to appear in the data).
To re-arrange and see the name used the most on top, add the `desc()` function:

```{r}
babynames_data %>% arrange(desc(n))
```

You'll see the name used most in one year is Linda in 1947 -- there were 99,686 female Lindas born that year.
The next function we'll try out is `filter()`, which allows you to only view certain rows of your data based on some criteria you set.
So, for example, if you only want to view the information for one name:

```{r}
babynames_data %>% filter(name=="Evelyn")
```

OK, let's unpack a few things about that code: 1.
When analyzing data, your "queries" will often start this way: piping your dataset (in this case stored in the "babynames_data" variable) into a function or series of functions that transform your data.
2.
Inside the `filter()` function, we set the criteria for the rows we want returned: in this case, the value in name has to match exactly "Evelyn".
In R you must use two equals signs (==) for this.
3.
R is almost always case sensitive: when typing column names, matching exact values ("Evelyn", not "evelyn"), and even in function names.

*Your turn!* Filter for a name of your choice:

```{r}

```

You can filter across multiple column names by using `&` to separate your filtering criteria.

```{r}
babynames_data %>% filter(name=="Evelyn" & year==2017)
```

One thing to notice here is when filtering for the name Evelyn, we used quotation marks around our desired result.
We did not do that when filtering for our desired year, 2017.
This is because the `name` column data type is character, while the `year` column is numeric.

*Your turn!* Look up how many babies were born with your name in your birth year:

```{r}

```

Next, let's string two functions together.
You can pipe your dataset through two (or more!) functions.
Keep in mind that this works linearly: the first function will do its work, and the output is piped into the next function.
Let's find out what the most popular baby names were in 2017.
First, pipe your dataset into the `filter()` function, then pipe it into the `arrange(desc())` functions:

```{r}
babynames_data %>% filter(year==2017) %>% arrange(desc(n))
```

Now that we've practiced sorting and filtering, let's turn to the third pillar of data analysis: aggregating.
In R we use a lot of the same summary functions as spreadsheets: `sum()`, `mean()`, `median()`.
R uses "mean" instead of "average." These functions are part of base R, so they don't inherently work with tidyverse's piping scheme, so we use the tidyverse `summarise()` function to receive the output and work with the summary functions.
Let's calculate the total number of baby names in our dataset.
Use `summarise()` and `sum()` together, and we'll give the new aggregated data the name `total_names`:

```{r}
babynames_data %>% summarise(total_names = sum(n))
```

Another common summary statistics we use in data analysis is the count of values, or the number of rows in a table.
The function for counting rows in R is `n()`, and doesn't require any arguments:

```{r}
babynames_data %>% summarise(num_rows = n())
```

This isn't super helpful for our babynames data, as many names are repeated across the table since they are used from year to year.
But we can string together two functions and determine, for example, how many unique names were given in 2017:

```{r}
babynames_data %>% filter(year==2017) %>% summarise(unique_names_2017 = n())
```

Suppose we want to see how many unique names were used across all years, not just 2017.
If you use spreadsheets much, hopefully what came to mind was: a pivot table!
We want to group our data based on whatever value is in the `year` column, and then do some summaries on those groups.
The function for this is `group_by()`.
Note: if you only use the `group_by()` function, R will create groups in the background of your data, but the output won't look any different:

```{r}
babynames_data %>% group_by(year)
```

But notice that next to the box that says "A tibble: 1,924,665 x 5" is a box that says "Groups: year [138]".
It has done the grouping in the background, and there are 138 groups (138 years; 1880 to 2017).
Pipe this into a `summarise()` function and it will act on the groups rather than the whole dataset.

```{r}
babynames_data %>% group_by(year) %>% summarise(unique_names = n())
```

To resort our results and bring the year with the most unique names to the top, add `arrange(desc())` to your code:

```{r}
babynames_data %>% group_by(year) %>% summarise(unique_names = n()) %>% arrange(desc(unique_names))
```

*Your turn!* Taking all that you know about piping and functions so far, can you find the name given to the most female babies across all years?

```{r}

```
