---
title: "Example analysis"
author: "Christian McDonald"
output:
  html_document:
    df_print: paged
---

## The whole game: Denied

This is an example of data cleaning, analysis and charting in a single pipeline using an R Notebook, based on a real investigative analysis.

Our goal here is to see how the rate of students in special education has changed over time in Texas. The code below has very light explanation, but you can see more detail in the "Denied" chapters in [Reporting with Data in R](https://utdata.github.io/rwdir/denied-cleaning.html), which follows up on a 2016 series from the Houston Chronicle.

A little background: The Houston Chronicle found that the Texas Education Agency in 2004 "arbitrarily decided what percentage of students should get special education services — 8.5% — and then forced school districts to comply by strictly auditing those serving too many kids."

Following that reporting by the Chronicle and others, the Texas Legislature in 2017 unanimously banned using a target or benchmark on how many students a district or charter school enrolls in special education.

We want to find if the percentage of special education students has changed since that time.

To do this analysis we have to import multiple years of data and bind them, join that with a reference table, do some filtering, summarize some values and then chart the result.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
```

## Importing multiple files at once

While a little more complicated than simply importing a file, we tackle a common challenge: Importing multiple files from a directory.

### Build a list of files

We have special education rates per district for an 10-year span. We want to build a "list" of the file names, then iterate over that list to import and join them.

```{r}
dstud_files <- list.files(
  "../data",
  pattern = "DSTUD",
  full.names = TRUE
)

dstud_files
```

### Read and combine

Now we capture the names of the files with `set_names()` and then use `map()` to apply the `read_csv()` function to build a list of data frames from each file. We then use `list_bind()` to put them together and `clean_names()` to normalize the variables/column names.

I slice the "head" and "tail" of the combine file to show there is a difference in the `district` column where the `'` was added to ensure the ID is a text value.

```{r}
dstud_raw <- dstud_files |>
  set_names(basename) |>
  map(
    read_csv,
    col_select = c(DISTRICT, DPETALLC, DPETSPEC, DPETSPEP)
  ) |> 
  list_rbind(names_to = "source") |>
  clean_names()

dstud_raw |> slice_head(n = 5)
dstud_raw |> slice_tail(n = 5)
```

### Some cleanup of columns

Now to clean up that district column and use the file name in our source column to build a year of the record.

```{r}
dstud <- dstud_raw |>
  mutate(
    district = str_remove(district, "'"), # removes the apostrophe
    year = str_c("20", str_sub(source, 7, 8)) # Create year from file name
  ) |> 
  select(!source) # remove old column

dstud |> glimpse()
```

## Joining files together

Our data does not have district names, but we'll fix that by joining with a reference file.

### Import our reference file

As we import this file using `read_csv()` we feed it the argument of `col_select = ` allowing us to select specific files. Somehow you have to know what that list is, so you usually add this after you've imported and inspected the entire file.

```{r}
dref_raw <- read_csv(
  "../data/DREF_22.csv",
  col_select = c(
    DISTRICT,
    CNTYNAME,
    DISTNAME,
    DFLCHART,
    DFLALTED
    )
  ) |> 
  clean_names()

dref_raw |> slice(1:5)
```

Select columns of interest and fix the district value to remove apostrophe.

```{r}
dref <- dref_raw |> 
  select(district, distname, cntyname, dflalted, dflchart) |> 
  mutate(district = str_remove(district, "'"))

dref |> slice(1:5)
```

### Join the files together

Here we start with the reference file so those columns are shown first, then do an `inner_join()` to join the student data, using the district column as our key.

```{r}
sped_joined <- dref |>
  inner_join(dstud, join_by(district))

sped_joined |> slice(1:5)
```

### Filter and focus

We filter out alternative education districts and then select our main columns of interest while also renaming some of them to be more understandable.

```{r}
sped <- sped_joined |> 
  filter(dflalted == "N") |> 
  select(
    district,
    distname,
    charter = dflchart,
    cntyname,
    year,
    all_count = dpetallc,
    sped_count = dpetspec,
    sped_percent = dpetspep
  )

sped |> slice(1:5)
```

## Summarize statewide data

We will use the count of all student and special education students in these districts to create a state-wide rate for traditional public schools.

```{r}
yearly_percent <- sped |> 
  group_by(year) |> 
  summarise(
    total_students = sum(all_count),
    total_sped = sum(sped_count)
  ) |> 
  mutate(sped_percent = ((total_sped / total_students) * 100) |> round(1))

yearly_percent
```

### Plot statewide data

Now we plot that data as a column chart.

```{r}
yearly_percent |>
  ggplot(aes(x = year, y = sped_percent)) +
  geom_col(fill = "cadetblue") +
  geom_text(aes(label = sped_percent, vjust = -.5)) +
  ylim(0,13) +
  labs(
    title = "Special education rates rise after rule change",
    subtitle = str_wrap("After 2016 reports highlighting a Texas Education Agency benchmark for special education, the Texas Legislature banned the practice and increased funding for programs."),
    x = "", y = "% of students in special education",
    caption = "By Christian McDonald; Source: Texas Education Agency"
  )
```

### Plot local districts

Let's look at some individual districts in the Austin area to make sure the trend tracks. There isn't any summarization needed to do this ... we just filter our combined data for districts of interest and then plot them.

Filtering here is for display purposes. We exclude one district that is an online-only school that serves students outside the area, and charter schools because there are too many to show effectively on one chart.

```{r}
sped |> 
  filter(
    charter != "Y",
    cntyname == "TRAVIS",
    distname != "UNIVERSITY OF TEXAS AT AUSTIN H S"
  ) |> 
  ggplot(aes(x = year, y = sped_percent, group = distname)) +
  geom_line(aes(color = distname)) +
  labs(
    title = "Special education rates for Travis County area school districts",
    x = "", y = "% in special education", color = "Districts"
  ) +
  ylim(0,16)
```



